Seminar 9: Cluster Analysis and PCA
========================================================
by Erica Acton
```{r include=FALSE}
library(lattice)
library(knitr)
library(RColorBrewer)
library(cluster)
library(pvclust)
library(xtable)
library(limma)
library(plyr)
```
Load photoRec dataset.
```{r }
prDat <- read.table("/home/eacton/R/stat540-2014-ACTON-ERICA/GSE4051_data.tsv", header=TRUE, row.names=1)
str(prDat, max.level=0)
prDes <- readRDS("/home/eacton/R/stat540-2014-ACTON-ERICA/GSE4051_design.rds")
str(prDes)
```
Rescale the rows.
```{r}
sprDat <- (t(scale(t(prDat))))
str(sprDat, max.level=0, give.attr=FALSE)

round(data.frame(avgBefore=rowMeans(head(prDat)),
                 avgAfter=rowMeans(head(sprDat)),
                 varBefore=apply(head(prDat),1,var),
                 varAfter=apply(head(sprDat),1,var)),2)
```
Hierarchical Clustering
=======================================================================

Computer pairwise distances.
```{r}
pr.dis <- dist(t(sprDat), method="euclidian")
```
Create a new factor representing the interation of gType and devStage.
```{r}
prDes$grp <- with(prDes, interaction(gType, devStage))
summary(prDes$grp)
```
Compute hierarchical clustering using different linkage types.
```{r}
pr.hc.s <- hclust(pr.dis, method='single')
pr.hc.c <- hclust(pr.dis, method='complete')
pr.hc.a <- hclust(pr.dis, method='average')
pr.hc.w <- hclust(pr.dis, method='ward')
```
Plot the different hierarchical clustering types.
```{r}
op <- par(mar=c(0,4,4,2),mfrow=c(2,2))

plot(pr.hc.s, labels=FALSE, main="Single", xlab="")
plot(pr.hc.c, labels=FALSE, main="Complete", xlab="")
plot(pr.hc.a, labels=FALSE, main="Average", xlab="")
plot(pr.hc.w, labels=FALSE, main="Ward", xlab="")

par(op)
```
Identify 10 clusters.
```{r}
op <- par(mar=c(1,4,4,1))

plot(pr.hc.w, labels=prDes$grp, cex=0.6, main="Ward showing 10 Clusters")
rect.hclust(pr.hc.w, k=10)

par(op)
```
Heatmap example.
```{r}
GreyFun <- colorRampPalette(brewer.pal(n=9, "Greys"))
gTypeCols <- brewer.pal(n=11, "RdGy")[c(4,7)]
heatmap(as.matrix(sprDat), Rowv=NA, col=GreyFun (256), hclustfun= function (x) hclust(x, method='ward'),
        scale="none", labCol=prDes$grp, labRow=NA, margins=c(8,1), ColSideColor=gTypeCols[unclass(prDes$gType)])
legend("topright", legend=levels(prDes$gType), col=gTypeCols, lty=1, lwd=5, cex=0.5)
```
Exercise: Playing with heatmaps.
----------------------------------------------------------------------------
```{r}
GnBuFun <- colorRampPalette(brewer.pal(n=9, "GnBu"))
gTypeCols <- brewer.pal(n=9, "RdGy")[c(4,7)]
heatmap(as.matrix(sprDat), Rowv=NA, col=GnBuFun (256), hclustfun= function (x) hclust(x, method='average'),
        scale="none", labCol=prDes$grp, labRow=NA, margins=c(8,1), ColSideColor=gTypeCols[unclass(prDes$gType)])
legend("topright", legend=levels(prDes$gType), col=gTypeCols, lty=1, lwd=5, cex=0.5)
```
K-means Clustering
---------------------------------------------------------------------------
Choose parameters, including k.
```{r}
set.seed(31)
k <- 5
pr.km <- kmeans(t(sprDat), centers=k, nstart=50)
```
Look at the sum of squares of each cluster.
```{r}
pr.km$withinss
```
Look at the composition of each cluster.
```{r}
pr.kmTable <- data.frame(devStage=prDes$devStage, cluster=pr.km$cluster)
prTable <- xtable(with(pr.kmTable, table(devStage, cluster)), 
                  caption='Number of samples from each developmental stage within each k-means cluster')

align(prTable) <- "lccccc"
print(prTable, type='html', caption.placement='top')
```
PAM Algorithm
------------------------------------------------------------------------------
```{r}
pr.pam <- pam(pr.dis, k=k)
pr.pamTable <- data.frame(devStage=prDes$devStage, cluster=pr.pam$clustering)
pamTable <- xtable(with(pr.pamTable, table(devStage, cluster)), caption='Number of samples from each developmental stage within each PAM cluster')

align(pamTable) <- "lccccc"
print(pamTable, type='html', caption.placement='top')

summary(pr.pam)
```
Silhouette plot.
```{r}
op <- par(mar=c(5,1,4,4))
plot(pr.pam, main="Silhouette Plot for 5 Clusters")
par(op)
```
Gene Clustering
----------------------------------------------------------------------------
Start with the top 972 genes that showed differential expression across the different developmental stage (BH adjusted p value < 10-5).
```{r}
devDes <- model.matrix(~devStage, prDes)
fit <- lmFit(prDat, devDes)
ebFit <- eBayes(fit)

topDat <- topTable(ebFit, coef = grep("devStage", colnames(coef(ebFit))), p.value=1e-05, n=972)

ttopDat <- sprDat[rownames(topDat), ]
head(ttopDat)
```
Hierarchical:
```{r}
geneC.dis <- dist(ttopDat, method='euclidean')
geneC.hc.a <- hclust(geneC.dis, method='average')

plot(geneC.hc.a, labels=FALSE, main="Hierarchical with Average Linkage", xlab="")
```
Partitioning:
```{r}
set.seed(1234)
k <- 5
kmeans.genes <- kmeans(ttopDat, centers=k)
```
Choose desired cluster.
```{r}
clusterNum <- 1
```
Set up axes.
Plot the expression of all the genes in the selected cluster in grey.
Add in the cluster center.
Colour points to show the developmental stage.
```{r}
plot(kmeans.genes$centers[clusterNum,], ylim=c(-4,4), type='n',xlab="Samples", ylab="Relative Expression")

matlines(y=t(ttopDat[kmeans.genes$cluster==clusterNum, ]), col='grey')
points(kmeans.genes$centers[clusterNum,],type='l')
points(kmeans.genes$centers[clusterNum,], col=prDes$devStage,pch=20)
```
Heatmaps (hierarchical):
```{r}
devStageCols <- brewer.pal(n=11, "RdGy")[c(2,4,7,9,11)]
heatmap(as.matrix(ttopDat), col=GreyFun (256), hclustfun= function (x) hclust(x, method='average'), scale="none", labCol=prDes$grp, labRow=NA, margins=c(8,1), ColSideColor=devStageCols[unclass(prDes$devStage)])
legend("topleft", levels(prDes$devStage), col=devStageCols, lty=1, lwd=5, cex=0.5)
```
Redefining the Attributes
-------------------------------------------------------------------------------
Define new attributes for a gene and estimate the parameters.
```{r}
annoTopDat <- stack(as.data.frame(ttopDat))
annoTopDat$probeset <- rownames(ttopDat)

annoTopDat <- merge(annoTopDat,prDes,by.x="ind", by.y="sidChar")
devStageAvg <- ddply(annoTopDat, ~probeset, function(x) {
  avgbyDevStage <- aggregate(values ~ devStage,x,mean)$values 
  names(avgbyDevStage) <- levels(x$devStage) 
  avgbyDevStage
  })

rownames(devStageAvg) <- devStageAvg$probeset
devStageAvg$probeset <- NULL
str(devStageAvg)
```
Make a heatmap.
```{r}
heatmap(as.matrix(devStageAvg), Colv=NA, col=GreyFun(256), hclustfun=function(x) hclust(x, method="average"),labCol=colnames(devStageAvg), labRow=NA, margin=c(8,1))
```
Look at the relative expression in all clusters with respect to developmental stage as determined by kmeans.
```{r}
k <- 4
geneDS.km <- kmeans(devStageAvg, centers=k, nstart=50)
clust.centers <- geneDS.km$centers
```
Plot all cluster centers separately.
```{r}
op <- par(mfrow=c(2,2))
for(clusterNum in 1:4) {
  plot(clust.centers[clusterNum,], ylim=c(-4,4), type='n', 
       xlab="Developmental Stage", ylab="Relative Expression", axes=F, 
       main=paste("Cluster", clusterNum, sep=""))
  axis(2)
  axis(1,1:5, c(colnames(clust.centers)[1:4], "4W"), cex.axis=0.9)

  matlines(y=t(devStageAvg[geneDS.km$cluster==clusterNum,]), col='grey')
  points(clust.centers[clusterNum,],type='l')
  points(clust.centers[clusterNum,], pch=20)
  }
par(op)
```
Plot to compare all clusters' centers.
```{r}
plot(clust.centers[clusterNum,], ylim=c(-4,4), type='n', 
       xlab="Developmental Stage", ylab="Relative Expression", axes=F, 
       main="Clusters' Centers")
  axis(2)
  axis(1,1:5, c(colnames(clust.centers)[1:4], "4W"), cex.axis=0.9)

for(clusterNum in 1:4) {
  points(clust.centers[clusterNum,],type='l', col=clusterNum, lwd=2)
  points(clust.centers[clusterNum,], col=clusterNum, pch=20)
}
```
Plotting 3-dimensional clusters as determined by k-means.
```{r}
cloud(devStageAvg[,"E16"]~devStageAvg[,"P6"]*devStageAvg[,"4_weeks"], col=geneDS.km$clust, xlab="E16", ylab="P6", zlab="4_weeks")
```
Statistical Measures to Evaluate Clusters
--------------------------------------------------------------------------
Qualifying cluster membership.
```{r}
pvc <- pvclust(ttopDat, nboot=100)
plot(pvc, labels=prDes$grp, cex=0.6)
pvrect(pvc, alpha=0.95)
```
PCA (Principal Components Analysis)
---------------------------------------------------------------------------
```{r}
#scree plot
pcs <- prcomp(sprDat, center=F, scale=F)
plot(pcs)
#scatterplot relating PCs to covariates
prinComp <- cbind(prDes, pcs$rotation[prDes$sidNum,1:10])
plot(prinComp[,c("sidNum", "devStage","gType","PC1","PC2","PC3")], pch=19, cex=0.8)
#plot data on 2 PCs, coloured by devStage
plot(prinComp[,c("PC1","PC2")], bg=prDes$devStage, pch=21, cex=1.5)
legend(list(x=0.2, y=0.3), as.character(levels(prDes$devStage)), pch=21, pt.bg=c(1,2,3,4,5))
```
